Cancer Applications of Choropleth maps, and the Potential of Cartograms and Alternative Map Displays

### 2019-12-12 review_article.Rmd
editing to camparisons of alternative displays

---
title: "Cancer Applications of Choropleth maps, and the Potential of Cartograms and Alternative Map Displays"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  word_document:
    toc: no
    toc_depth: '3'
link-citations: yes
fontsize: 11pt
csl: vancouver.csl
yolo: yes
---

@@

Cancer atlases communicate cancer statistics over geographic domains. These domains are subdivided into administrative regions such as countries, states or suburbs. When communicating human-related statistics in Australia, the geographic map base draws attention to the sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers but they are important to consider if they are more densely populated. Alternative map displays can reveal patterns that are not obvious when the same data is shown on a choropleth map. Alternative displays have proven to be successful on online news sites, especially for the communication of election data, where votes are aggregated at state level.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Cancer statistics are often delivered as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations into statistics for the geographical units, especially for privacy protection but also for political and policy purposes. Counts per area require minimal transformation (e.g. state, province, local government area, post/zip code). However, counts alone are not sufficient for the comparison areas, because the populations of areas are all different. The count data needs to be merged with population data to present incidence rates, for example, the rate per 100000 people. This type of data is collected on a routine basis for public health purposes and may be made available to the general public as a service to the community.

<!-- Ways to visualize, whats our hypothesis -->
Map displays are commonly required for the communication of cancer statistics over geographic domains. When using a choropleth map: the statistic is mapped to color and the geographic region is filled with this color. Disease maps help to present geographic patterns that may be overlooked in a table, obscuring the geospatially related statistics [@SAMGIS]. Providing a visual representation of cancer outcomes allows geographic patterns of the disease to be identified and effectively addressed with public health policy and actions. The spatial distribution of the disease incidence can be examined using a choropleth and may reveal a trend in longitude or latitude, or rural vs urban, or coastal vs inland, or even specific hot spots of the disease. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and effectively helps in the formulation of hypotheses. 
Daniel Exeter [@SE] recognizes one of the key challenges with mapping spatial patterns of disease is the design of visualizations. This paper considers the current visualization techniques to communicate statistics to the public, and their applications to cancer statistics. Alternative approaches are also considered, and the limitations of the visualization methods, highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is organized as follows. The next section describes the choropleth map which is the common approach to disease maps and presents examples of atlases in use today. The limitations of the choropleth map are discussed in Section \ref{ch:chorolimit}. Section \ref{ch:alternatives} describes alternative displays, including the cartogram which is useful when the map has heterogeneously sized geographic units. Section \ref{sec:compare} presents the limitaions of producing and using alternative displays. Disease maps are more useful when made interactive, and common options are described in Section \ref{ch:interacting}, along with a discussion of benefits and disadvantages.

# Traditional approaches for cancer map displays {#ch:choropleths}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth map displays the geographical distribution of data over a set of spatial units by shading areas of a map [@EI]. The geography of a choropleth is faithfully rendered, and the color rendering is designed to reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS].
Early versions of choropleth maps used symbols or patterns instead of color.
Choropleths can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure \ref{fig:choroCRS} shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged from 2012 to 2016. The data was extracted from the official federal statistics on cancer incidence and deaths [@USInteractive], produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). The rates exhibit a spatial trend, increasing from west to east. There is also a spatial outlier -- Utah has a noticeably lower rate than its neighbors. Also, Kentucky and Maine have higher rates than their neighbors. There is possibly a cluster of higher rates around the tobacco states.

<!-- psychology -->
Displaying familiar state boundaries can make a map simpler to read [@CIBMUK], and allow viewers to visually infer the spatial relationships in the data using their mental model of the geography.
Consumers of disease displays may include researchers, the general public, policymakers and the media [@CPISACA], the familiarity of the geography is a worthy consideration when presenting results of spatial analysis.

<!-- potential problems -->
The familiar shapes are often politically driven boundaries with individual areas being of non-uniform size, containing different population densities and subject to change over time. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the underpopulated but large areas [@EI], Skowronnek [@BCM] calls this an area-size bias. 
Choropleths can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Publicly available choropleth cancer maps..", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleths -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country, or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. 
Cancer maps and atlases date back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power, and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the general public and others not trained in statistical analyses.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analyses. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps utilized to communicate to the public and other non-expert audiences are choropleths.

The statistics presented have been developed over time. Mortality rates are now often presented as relative rates of risk across the population, and age-adjusted to correct for the higher prevalence of cancers in older populations.
Howe [@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table \ref{tab:measures} summarizes the measures commonly presented in published cancer atlases.

Table: \label{tab:measures} Commonly used measures for reporting cancer information.

@@

<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 publicly available cancer atlases, published between 2010 and 2018, from different countries around the globe. All of these online atlases used choropleth maps. All but one of these were published by non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships.  Figure \ref{fig:choropleth-grid-create} displays a subset of maps from these atlases.

The selection varies in the geographies explored, Figure \ref{fig:choropleth-grid-create}b, Globocan 2012 [@Globocan] explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide utilising data sourced from cancer registries of each country. 
The Bowel Cancer Australia Atlas presents an example of a cancer specific atlas, Figure \ref{fig:choropleth-grid-create}d shows the average Standardised Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia [@Bowel]. The choice of gender can be selected in the Bowel Cancer Atlas, an alternative approach in employed in the Map of Cancer Mortality Rates in Spain, Figure \ref{fig:choropleth-grid-create}f shows side by side maps of the relative risk of lung cancer separately for men and women [@cancerSpain].

There are vast differences in the geographies considered by online atlases. 
Figure \ref{fig:choropleth-grid-create}b shows global age-standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018.
The United States Cancer Statistics [@USInteractive] can be seen in Figure \ref{fig:choropleth-grid-create}e. This map contains the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level.
The Environment and Health Atlas of England and Wales, Figure \ref{fig:choropleth-grid-create}a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2] at a neighbourhood (small-area) scale.
The Atlas of Cancer in Queensland, Figure \ref{fig:choropleth-grid-create}c shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 in the state of Queensland within Australia [@QLDcancerAtlas].

Age specific atlases are less common, Figure \ref{fig:choropleth-grid-create}g displays Atlas of Childhood Cancer in Ontario, this communicates the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric]. This atlas communicates information regarding a very small geographic region.

## Additional materials
<!-- 479 words -->
Supplementary graphs and plots are often included to add more depth and information. Additional materials such as tables, graphs, and text explanations, support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, standard deviation. When presenting cancer maps, the intuition derived from maps must be 'validated by rigorous statistical analyses' [@MACM], and the display of the supplementary statistics helps in this regard.

### Geographic hierarchies

While atlases are often used to describe differences between areas, statistics may be displayed at different levels of aggregation. Global health statistics can be aggregated to administrative and arbitrarily defined regions, such as those used by the World Health Organization and the United Nations [@IARC_3].
World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USInteractive].

### Population distribution

Each population area will likely have a different number of people. The distribution of the population residing in all areas may also be communicated in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density. Alternatives to choropleth to address this are described in Section \ref{ch:alternatives} .

### Statistical uncertainty

Uncertainty about the value of a statistic is often communicated in cancer atlases. Uncertainty may occur for several reasons: numbers are drawn from samples, errors occur in the disease reporting, or values are simulated from a model to maintain privacy.  Uncertainty is often recorded as sample or population size, standard deviation, confidence intervals, statistical significance, quantiles, credible intervals. The most common measure used to represent uncertainty were credible or confidence intervals (CIs).

Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it can be difficult to display effectively. The map focuses on displaying the statistic and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two approaches.

### Demographics

Demographics include information regarding the age and sex distribution of the areas presented. Digital atlases allow for users to control the display, to select subsets such as males, females or those aged over 65.

### Socio-economic indicators

Socio-economic indicators, such as unemployment rates, poverty rates, remoteness, and education levels,  can explain how the experience of cancer prevalence varies for various members of society. Few atlases provide this level of detail. Data is available from the United Nations Human Development Index and financial indicators are available from the World Bank [@IARC_3].

@@

A choropleth display may be limited by the administrative boundaries used to define regions.
This issue is often discussed in the context of election mapping, where each electorate decides the outcome of one parliamentary seat in the house or representatives.
While all states are visible in a map display of Australia, smaller regions such as electorates present an extreme case of an urban-rural divide. 
The landmass occupied by urban electoral districts is only 10% of Australia, yet 90% of the population live in these urban areas. To present election results on a choropleth map should be 'unthinkable', as it means diminishing the visual impact of the majority of the electorates.
Using a map of Australia as a visual statistic would communicate an incorrect distribution of votes across the country. The issue of spatially heterogeneous population distribution across a country's landmass is common to many countries, especially British colonies [@ACTUC].

Choropleths provide a familiar display. This may be less threatening to an audience that could be frightened by statistics, but they suffer primarily from unfaithfully representing the disease information by hiding small geographic areas. Inset maps like in Brisbane city in Figure \ref{fig:choropleth-grid-create}c of the state of Queensland, are commonly used to reduce distorted interpretations, but it is a band aid remedy.

 As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.

# Contemporary alternatives to choropleths {#ch:alternatives}

@@

<!-- What is a cartogram? -->
Choropleths imply uniformity of data across the geographic space however population densities are extremely unlikely to be uniform [@BCM]. The cartogram was developed to address this, by transforming the map to better indicate population [@ACCAC]. The resulting display more accurately communicates the impact of the disease across the population, as recorded by the statistic, at the sacrifice of geographic accuracy.

<!-- There are many alternative views of the world-->
<!-- Tobler map projections - I think this is not relevant here
Choropleths may be considered true topological maps however,, if the landmass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of landmass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection. 
-->

@@

<!-- Denominators -->
The purposeful distortion of the map space, transformed according to population density, is beneficial when a uniform density of the map base is desired. The population then becomes a uniformly distributed background for the statistic presented [@GOINO], "population distribution is often extremely uneven in former British colonies" [@ACTUC], and this makes the distortion necessary [@CTTMB].
When implementing a distortion of the geographical shape according to population, the resulting display is an area cartogram [@NAC], or population-by-area cartogram [@TAAM]. Event cartograms change the area of regions on a map depending on the amount of disease-related events, but this does not consider the effects of land area and population [@VSSDCUC].

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape or distance of geographical areas [@ACTUC]. 
Monmonier [@HTLWM] suggests that white lies may be employed to create useful displays and map creators can draw lines that may distort the geometry and suppress features and it is easy for the average person to disregard the impact of transformations used to create cartograms.
Cartograms may be seen as an extension of map transformations and projections. The favorable distortion is proportional to a value other than the actual earth size area [@NAC].
A disadvantage of the conventional map is that sparsely populated rural areas may be emphasized, whereas the areas representing cities are very small, interpreting spatial patterns very difficult.
The distortion of a cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC].
When visualizing population statistics, Dorling [@ACTUC] considers this equitable representation design 'more socially just', or honest [@NISCC], giving due attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas [@CTTMB]. 
<!-- @MTMSIH encourage the use of cartograms to highlight small areas and uncover local-level inequalities. DI SAYS I think that focuses on Bayesian analysis not on cartograms-->

@@

The creation of cartograms was historically in the hands of professional cartographers [@CD]. 
Early approaches including John Hunter and Jonathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball-bearing approach and Tobler's (1973) computer programs [@ACTUC]. 
Geographical information systems allowed map users, and researchers to create cartograms, but these systems are utilized depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994), [@CD].
Howe [@HEDP] discusses the impact of electronic computer-assisted techniques.

There are many alternatives to consider, the intended audience of the map, and its purpose are key points in cartogram use and creation.
 'There is no "best" cartogram or method of creating cartograms just as there is no "best" map' (Monmonier and Schnell, 1988) [@ACTUC]. 
Nusrat and Kobourov [@SAIC] provided a framework to investigate implementations of the many algorithms presented, and the "statistical accuracy, geographical accuracy, and topological accuracy".

>>>>

---
title: "Cancer Applications of Choropleth maps, and the Potential of Cartograms and Alternative Map Displays"
author: "Stephanie Kobakian, Jessie Roberts and Dianne Cook"
bibliography: reviewbib.bib
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: no
    toc_depth: 3
  word_document:
    toc: no
    toc_depth: '3'
link-citations: yes
fontsize: 11pt
csl: springer-vancouver.csl
yolo: yes
---

@@

Cancer atlases communicate cancer statistics over geographic domains. They subdivide these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics in Australia, the geographic map base draws attention to sparsely populated rural areas. The smaller geographic areas may not draw the attention of readers, but they are important to consider if they are more densely populated. Alternative map displays can be more effective than a choropleth map at revealing spatial patterns. Alternative displays are effective for the communication of election data aggregated at the state or electorate level on online news sites.

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments and not-for-profits are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations into the statistics for the geographical units, especially for privacy protection, but also for political and policy purposes.
Data collection or aggregation regions for the amount of cancer diagnoses or excess deaths per area include states, provinces, local government areas, and post/zip codes. To provide the amount, or count, of the diagnoses or excess deaths with these areas, it needs minimal transformation. However, the amount of people living in each of the areas in a set will be different. Those who wish to calculate rates can use the amount of people living in an area as a denominator and the diagnoses can become a numerator to create rates per amount of people. Health registries routinely collect data for public health reasons and may make this available to the public as a service to the community.

<!-- Ways to visualize, whats our hypothesis -->
Choropleth map displays are commonly used for the communication of cancer statistics over geographic domains. In this display, map creators map a statistic to color and apply the colour to the polygons representing geographic regions. Map users may overlook geographic patterns of disease when obscured in a table [@SAMGIS]. Visualizing diseases on maps is often the first step in exploratory spatial data analysis and helps in the formulation of hypotheses. Providing a visual representation of cancer outcomes allows identification, and public health policy and actions can address these patterns. choropleth maps allow the examination of the spatial distribution of the disease incidence, this may reveal a trend in longitude or latitude, differences between rural areas and urban, or coastal and inland areas, or even specific hot spots of the disease. Daniel Exeter [@SE] recognizes that one of the key challenges with mapping spatial patterns of disease is the design of visualizations. This paper considers the current visualization techniques to communicate statistics to members of the public and their applications to cancer statistics. It also considers alternative approaches and compares the visualization methods, highlighting the differences and historic use of these displays.

<!-- Structure of the paper -->
The paper is structured as follows. The next section describes the choropleth map, which is the common approach to disease maps and presents examples of atlases in use today. Section 3 discusses the limitations of the choropleth map. Section 4 describes alternative displays, including the cartogram, which is useful when the map has heterogeneously sized geographic units. Section 5 presents the limitations of the production and use of alternative displays. Disease maps are more useful when made interactive, Section 6 describes common options, along with a discussion of benefits and disadvantages.

# Traditional approaches for cancer map displays {#ch:choropleth maps}
<!-- 381 words -->
<!-- what is a choropleth map -->
A choropleth is traditionally used to make a cancer map. It is used to show the geographical distribution of data values by shading areas of a map [@EI]. A choropleth map involves faithfully rendering the geography, and the design of the color scheme and rendering reveal spatial patterns among data values. Identifying and explaining spatial structures, patterns, and processes involve considering the individuals and organizing them into representable units of communities [@SAMGIS]. Early versions of choropleth maps used symbols or patterns instead of color. choropleth maps can be used for displaying disease data [@DMAHP], including cancer data [@CPISACA]. In epidemiology, choropleth maps are often used as a tool to study the spatial distribution of cancer incidence and mortality.

Figure 2<!--\ref{fig:choroCRS}--> shows various geographic projections of choropleth maps of age-adjusted rate (per 100,000 people) of new cases of lung and bronchus in the USA, averaged from 2012 to 2016. The official federal statistics provide cancer incidence and deaths [5], produced by the Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI). The rates exhibit a spatial trend, increasing from west to east. Utah is a spatial outlier, it has a noticeably lower rate than its neighbors. Also, Kentucky and Maine have higher rates than their neighbors. There is possibly a cluster of higher rates around the tobacco states.

<!-- psychology -->
Displaying familiar state boundaries can make a map easier to read [@CIBMUK] and allow viewers to infer the spatial relationships visually in the data using their mental model of the geography.
Consumers of disease displays may include researchers, members of the public, policymakers, and the media  [@CPISACA]. For these consumers, the familiarity of the geography is a worthy consideration when presenting results of spatial analysis.

<!-- potential problems -->
Politically driven boundaries produced the now familiar shapes or regions, this has created individual areas of non-uniform size. These areas have different population densities and are subject to change. 
The different population and geographical sizes of administrative areas can attract attention to the shades of the underpopulated but large areas [@EI], Skowronnek [@BCM] calls this an area-size bias. 
choropleth maps can inhibit visual inference when presenting human-related statistics as the display may draw attention away from the 'potentially more important results in the more populous communities' that are geographically smaller [@SE].

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Examples of publicly available choropleth cancer maps.", out.width='100%', fig.align='center', fig.width=12, fig.height=10}
library(cowplot)

@@

## Cancer atlases {#ch:public}
<!-- 360 words -->
<!-- Displaying of cancer data on choropleth maps -->
A cancer atlas is a map, or collection of maps, representing cancer incidence and mortality for a country or group of countries. Atlases are key to developing hypotheses regarding areas with unusually high rates, geographic correlations, work-related exposures, and high-risk diets [@MACM]. The data collection methods across regions and the administrative control within regions lends itself to choropleth visualization. 
Cancer maps and atlases dating back to Haviland's maps in 1875, and early work in US cancer atlases appearing in 1971 [@burbank]. The presentation of cancer statistics has increased with greater access to computational power and the availability of geographic information systems software [@SE].

<!-- Measures reported in cancer atlases -->
Cancer maps are effective tools for communicating incidence, survival, and mortality to a wide range of audiences, including the general public and others not trained in statistical analysis.
These visualizations enable non-expert audiences to interpret the outputs of sophisticated statistical analysis. Cruickshank (1947) as cited by S. D. Walter [@DMAHP], discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly, the cancer maps used to communicate to members of the public and other non-expert audiences are choropleth maps.

Epidemiologists and statisticians have developed these statistics over several decades. Mortality rates commonly presented as relative rates of risk across the population and age-adjusted to correct for the higher prevalence of cancers in older populations.
Howe [@HEDP] describes Stock's development of the standardized mortality ratios through the 1930s. Table 1<!--\ref{tab:measures}--> summarizes the measures commonly presented in published cancer atlases.

Table: 1<!--\label{tab:measures}--> Common measures for reporting cancer information.

@@

<!-- 416 words -->
Roberts [@roberts2019communication] identified 33 publicly available cancer atlases, published between 2010 and 2018, from different countries around the globe. These online atlases used choropleth maps. Non-commercial organizations, including not-for-profits, government, research organizations, advocacy groups or government-funded partnerships published all but one of these. Figure 1<!--\ref{fig:choropleth-grid-create}--> displays a subset of maps from these atlases.

The selection varies in the geographies explored, Figure 1<!--\ref{fig:choropleth-grid-create}-->b, Globocan 2012 [@Globocan] explores Estimated Cancer Incidence, Mortality and Prevalence Worldwide utilizing data sourced from cancer registries of each country. 
The Bowel Cancer Australia Atlas presents an example of a cancer specific atlas, Figure 1<!--\ref{fig:choropleth-grid-create}-->d shows the average Standardized Incidence Ratio of colorectal cancer for Australian males from 2006 to 2010 in Australia [@Bowel]. Like many of the atlases examined, there is a choice of gender displayed in the Bowel Cancer Atlas. An alternative approach used in the Map of Cancer Mortality Rates in Spain, Figure 1<!--\ref{fig:choropleth-grid-create}-->f shows side-by-side maps of the relative risk of lung cancer separately for men and women [@cancerSpain].

There are vast differences in the geographies considered by online atlases. 
Figure 1<!--\ref{fig:choropleth-grid-create}-->b shows global age-standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018.
The United States Cancer Statistics [@USInteractive] presented in Figure 1<!--\ref{fig:choropleth-grid-create}-->e contains the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level.
The Environment and Health Atlas of England and Wales, Figure 1<!--\ref{fig:choropleth-grid-create}-->a shows the relative risk for women developing lung cancer in England and Wales in 2010 [@EnvEnglandWales2] at a neighborhood (small-area) scale.
The Atlas of Cancer in Queensland, Figure 1<!--\ref{fig:choropleth-grid-create}-->c shows the relative incidence ratio of lung cancer in males for each Statistical Area at Level 2 in the state of Queensland within Australia [@QLDcancerAtlas].

Age-specific atlases are less common, Figure 1<!--\ref{fig:choropleth-grid-create}-->g displays Atlas of Childhood Cancer in Ontario, this communicates the incidence rate of childhood cancers per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004 [@OntarioPediatric]. This atlas communicates information regarding a tiny geographic region.

## Additional materials
<!-- 479 words -->
Including supplementary graphs and plots add more depth and information. Additional materials such as tables, graphs, and text explanations, support the understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, standard deviation. When presenting cancer maps the display of the supplementary statistics helps. The intuition derived from maps must be ‘validated by rigorous statistical analyses’ [@MACM].

The statistics communicated in atlases are often used to describe differences between areas. This can occur at different levels of aggregation. Aggregation of global health statistics occurs within administrative and arbitrarily defined regions, such as those used by the World Health Organization and the United Nations [@IARC_3].
World atlases can allow for displays of data aggregated into continents, countries, states, provinces and congressional districts [@USInteractive].
Each population area will probably have a different number of people. Cancer atlases may also communicate the distribution of the population living in all areas in a table or histogram display [@NICR_1].
Atlases can connect the population to the land available to them by communicating population density. Section  4<!--\ref{ch:alternatives}--> considers alternatives to choropleth that introduce population into the visualizations.

Map users can define the population further using the demographics of the individuals within each area.  Demographics that can influence cancer rates include information regarding the age and sex distribution of the areas presented. Some digital atlases surveyed allow users to control the display to select subsets such as males, females or those aged over 65. 
Similarly, socioeconomic indicators, such as unemployment rates, poverty rates, remoteness, and education levels,  can explain how the experience of cancer prevalence varies for various members of society. Few atlases provide this level of detail.
Introducing population and demographic information helps to interpret the rates in areas effectively, but there will still be uncertainty around the rates.
To address this, cancer atlases often communicate uncertainty about the value of a statistic. Cancer atlas creators may introduce uncertainty when using samples, or there are errors occurring in the disease reporting, or models are used to simulate values to maintain privacy. The statistics that report on uncertainty include sample or population size, standard deviation, confidence intervals, statistical significance, quantiles, and  credible intervals. The most common measures used to represent uncertainty were credible or confidence intervals (CIs). Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it is difficult to display effectively. The map focuses on displaying the statistic and lacks additional space to represent the uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two approaches.

@@

The administrative boundaries used to define regions may limit a choropleth display.
Those who critique choropleth maps often dicuss election mapping, where each electorate decides the outcome of one parliamentary seat in the house of representatives.
While all states are visible in a map display of Australia, smaller regions such as electorates present an urban-rural divide, where rural regions are much larger than metropolitan areas of similar or greater populations. 
The landmass occupied by urban electoral districts is only 10% of Australia, yet 90% of the population live in these urban areas. To present election results on a choropleth map should be 'unthinkable',XXX as it means diminishing the visual impact of most of the electorates.
Using a map of Australia as a visual statistic would communicate an incorrect distribution of votes across the country.

Spatially heterogeneous population distribution across a country's landmass is an issue common to many countries, especially British colonies [@ACTUC]. However, the boundaries shown in choropleth maps provide a familiar display. This may be less threatening to an audience that could be unfamiliar was interpreting spatial statistics, even though they suffer primarily from unfaithfully representing the disease information by hiding small geographic areas. Inset maps like in Brisbane city in Figure 1<!--\ref{fig:choropleth-grid-create}-->c of the state of Queensland are commonly used to reduce distorted interpretations, but it is a band aid remedy. 
As this feature of population distributions continues to intensify, the need for cartograms as an alternative to a choropleth map should only increase.

# Contemporary alternatives to choropleth maps {#ch:alternatives}

@@

<!-- What is a cartogram? -->
choropleth maps imply uniformity of data across the geographic space but population densities are unlikely to be uniform [@BCM]. Cartographers developed the cartogram to draw the attention to the population by transforming the map [@ACCAC]. The resulting display can communicate the impact of the disease more accurately across the population, as recorded by the statistic, at the sacrifice of geographic accuracy.

<!-- There are many alternative views of the world-->
<!-- Tobler map projections - I think this is not relevant here
choropleth maps may be considered true topological maps however,, if the landmass displayed covers enough of the globe, there must be a transformation or distortion to display the land in 2D [@HTLWM]. The amount of distortion is related to the distance covered by the landmass displayed @GAMP.
World map projections reflect the frequent perspectives used to view the earth. Choropleth maps will always be distorted if they cover enough of the globe, just like photographs of the globe from space. Choropleth creation requires choosing a map projection that shows a favorable distortion of the geography for presenting the set of spatial information. Selecting a display can prevent misinterpretation of global statistics, as global maps face the challenge of equitable displays of landmass on maps [@RSCW].
If the statistic presented on the map base relies on physical distance and is influenced by the topology there is no transformation needed, beyond choosing a reasonable projection. 
-->

@@

<!-- Denominators -->
When a map creator desires a uniform population density of the map base, the purposeful distortion of the map space is beneficial. The "population distribution is often extremely uneven in former British colonies" [@ACTUC], and this makes the distortion necessary, [@CTTMB] population will then become a uniformly distributed background for the statistic presented [@GOINO].
An area cartogram [@NAC], or population-by-area cartogram [@TAAM] occurs after the distortion of the geographical shape according to population. Event cartograms change the area of regions on a map depending on the amount of disease-related events, but this does not consider the effects of land area and population [@VSSDCUC].

<!-- Why transform?-->
<!-- Common variables used to create cartograms, e.g. population, mortality -->
Cartograms provide an alternative visualization method for statistical and geographical information.
The key difference between a choropleth and a cartogram is the desirable augmentation of the size, shape, or distance of geographical areas [@ACTUC]. 
Monmonier [@HTLWM] suggests that map creators can use white lies to create useful displays by distorting the geometry and suppressing features. It is easy for the average person to disregard the impact of transformations used to create cartograms. One disadvantage of the conventional map is interpreting spatial patterns is difficult because of the sparsely populated rural areas being emphasized in a choropleth map, and the inner city areas are geographically small.
The distortion of each area when making a cartogram is proportional to a value other than the actual earth size area [@NAC]. The distortion in an area cartogram accounts for the population density, preventing it from obscuring the spatial patterns [@TAAM]. 
The spatial transformation of map regions relative to the data emphasizes the data distribution instead of land size [@CBATCC]. When visualizing population statistics, Dorling [@ACTUC] considers this design 'more socially just', or honest [@NISCC], giving equitable representation and attention to all members of the population and reducing the visual impact of large areas with small populations [@DMAHP]. 
Howe [@HEDP] suggests that 'cancer occurs in people, not in geographical areas' and that spatial socio-economic data, like cancer rates, are best presented on a cartogram for urban areas as the population map base avoids allocating 'undue prominence' to rural areas [@CTTMB].

@@

The creation of cartograms was historically in the hands of professional cartographers [@CD]. 
Early approaches by John Hunter and Jonathan Young (1968) and Durham's wooden tile method, Skoda and Robertson's (1972) steel ball-bearing approach and Tobler's (1973) computer programs [@ACTUC]. 
Howe [@HEDP] discusses the impact of electronic computer-assisted techniques.
Geographical information systems allow map creators to produce cartograms and they use these systems depending on ‘the effectiveness, efficiency, and satisfaction of the map products' (Nielsen 1994) [@CD].

There are many issues to consider when creating alternative map displays: the intended audience of the map, and its purpose are key points in cartogram use and creation as 'there is no "best" cartogram or method of creating cartograms just as there is no "best" map' (Monmonier and Schnell, 1988) [@ACTUC]. 
Nusrat and Kobourov [@SAIC] provided a framework to investigate implementations of the many algorithms presented, and the "statistical accuracy, geographical accuracy, and topological accuracy".

```{r ggcartograms, fig.cap = "Common alternatives to maps, showing the same information as in Figure \ref{choroCRS} for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). Maps (a) - (c) are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help ease losing information about physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in (d) the hexagon tile map.", message=FALSE, warning=FALSE, out.width='100%'}
g1 <- png::readPNG("figures/ggcont.png")
g2 <- png::readPNG("figures/ggncont.png")
g3 <- png::readPNG("figures/ggdorl.png")
g4 <- png::readPNG("figures/gghexmap.png")
#knitr::include_graphics("figures/usa_grid.png")
